---
title: "downstream_analysis"
author: "Roy"
format: html
editor: visual
---

## Downstream Analysis

### Setup the working directory

```{r}
#| label: setup
#| include: false

knitr::opts_knit$set(
    root.dir = "/home/roy/Desktop/learning_projects/10X_Chromium_Data_Analysis"
)
      
```

### Install and load the necessary packages

```{bash}
# The command below is provided as an example for how to install r and biocoductor packages using mamba
mamba install -c conda-forge -c bioconda r-tidyverse bioconductor-seurat
```

```{r message=FALSE, warning=FALSE}

library(tidyverse)
library(Matrix)
library(SingleCellExperiment)
library(DropletUtils)
library(AcidPlots)
```

### Load the Data

#### Adipo 3d

```{r message=FALSE, warning=FALSE}

# loading the gene expression matrix
adipo_mat <- readMM(
    "data/processed-data/adipo_3d/counts_unfiltered/cells_x_genes.mtx"
)
#dim(adipo_mat)
#class(adipo_mat)

# loading the barcode and gene information
adipo_barcodes <- readLines(
    "data/processed-data/adipo_3d/counts_unfiltered/cells_x_genes.barcodes.txt"
)

adipo_genes <- readLines(
    "data/processed-data/adipo_3d/counts_unfiltered/cells_x_genes.genes.txt"
)

adipo_gene_names <- readLines(
    "data/processed-data/adipo_3d/counts_unfiltered/cells_x_genes.genes.names.txt"
)

```

Next we need to sanity check the data that is loaded into the environment.

```{r}
# Check the dimension and class of the gene expression matrix
dim(adipo_mat)
class(adipo_mat)

# check the lengths of the vectors containing information on barcodes, gene ID and names
length(adipo_barcodes)
length(adipo_genes)
length(adipo_gene_names)
```

The explanation of the output is given below:

788,172 barcodes –\> 788,172 rows in the original matrix

39,546 gene IDs –\> 39,546 columns in the original matrix

39,546 gene names –\> one name for every gene ID

```{r}
# transpose the gene expression matrix so that it is compatible to be turned into a SingleCellExperiment object

adipo_mat <- t(adipo_mat)

dim(adipo_mat)
```

Now we need to annotate the gene expression matrix with gene IDs and cell bar codes

```{r}
rownames(adipo_mat) <- adipo_genes
colnames(adipo_mat) <- adipo_barcodes

# check the matrix
adipo_mat[1:5, 1:5]
```

### Create the SingleCellExperiment object

```{r}
adipo_sce <- SingleCellExperiment(
    assays = list(
        counts = adipo_mat
    )
)

# adding the experimental condition
adipo_sce$condition <- "Adipo_3d"
```

kb count output —-\> Matrix Market file —-\> R sparse matrix —-\> SingleCellExperiment

```{r}
dim(adipo_mat)
length(adipo_barcodes)
length(adipo_genes)
dim(adipo_sce)
```

#### Cell calling using DropletUtils

```{r}
adipo_ranks <- barcodeRanks(counts(adipo_sce))
```

```{r fig.height=6, fig.width=8}
plot(
    adipo_ranks$rank,
    adipo_ranks$total,
    log = "xy",
    xlab = "Barcode rank",
    ylab = "Total UMI count",
    main = "Adipo 3d barcode rank"
)

o <- order(adipo_ranks$rank)

lines(
    adipo_ranks$rank[o],
    adipo_ranks$fitted[o]
)

abline(
    h = metadata(adipo_ranks)$knee,
    lty = 2
)

abline(
    h = metadata(adipo_ranks)$inflection,
    lty = 2
)
```

```{r}
metadata(adipo_ranks)$knee
metadata(adipo_ranks)$inflection
```

> ***Note to self: Need to better understand what exactly is happening in the bar code rank and how to properly interpret the plot and terminologies such as knee and inflection***

```{r}
# dropping droplets with ambient RNA (will take some time)

adipo_emptydrops <- emptyDrops(
    counts(adipo_sce)
)
```

```{r}
head(adipo_emptydrops)
```

```{r}
table(adipo_emptydrops$Limited)
```

```{r}
sum(adipo_emptydrops$FDR <= 0.01, na.rm = TRUE)
```
