---
title: "downstream_analysis"
author: "Roy"
format: html
editor: visual
---

## Downstream Analysis

### Setup the working directory

```{r}
#| label: setup
#| include: false

knitr::opts_knit$set(
    root.dir = "/home/roy/Desktop/learning_projects/10X_Chromium_Data_Analysis"
)
      
```

### Install and load the necessary packages

```{bash}
# The command below is provided as an example for how to install r and biocoductor packages using mamba

# mamba install -c conda-forge -c bioconda r-tidyverse bioconductor-seurat

# mamba install -c <channel-name> -c <second-channel-name> <repository>-<package>
```

```{r message=FALSE, warning=FALSE}

library(tidyverse)
library(Matrix)
library(SingleCellExperiment)
library(DropletUtils)
library(AcidPlots)
library(scuttle)
library(scDblFinder)
```

## Load the Data

### Adipo 3d

```{r message=FALSE, warning=FALSE}

# loading the gene expression matrix
adipo_mat <- readMM("data/processed-data/adipo_3d/counts_unfiltered/cells_x_genes.mtx")

# loading the barcode and gene information
adipo_barcodes <- readLines("data/processed-data/adipo_3d/counts_unfiltered/cells_x_genes.barcodes.txt")
adipo_genes <- readLines("data/processed-data/adipo_3d/counts_unfiltered/cells_x_genes.genes.txt")
adipo_gene_names <- readLines("data/processed-data/adipo_3d/counts_unfiltered/cells_x_genes.genes.names.txt")

```

Next we need to sanity check the data that is loaded into the environment.

```{r}
# Check the dimension and class of the gene expression matrix
dim(adipo_mat)
class(adipo_mat)

# check the lengths of the vectors containing information on barcodes, gene ID and names
length(adipo_barcodes)
length(adipo_genes)
length(adipo_gene_names)
```

The explanation of the output is given below:

788,172 barcodes –\> 788,172 rows in the original matrix

39,546 gene IDs –\> 39,546 columns in the original matrix

39,546 gene names –\> one name for every gene ID

```{r}
# transpose the gene expression matrix so that it is compatible to be turned into a SingleCellExperiment object
adipo_mat <- t(adipo_mat)

dim(adipo_mat)

```

Now we need to annotate the gene expression matrix with gene IDs and cell bar codes

```{r}
rownames(adipo_mat) <- adipo_genes
colnames(adipo_mat) <- adipo_barcodes

# check the matrix
adipo_mat[1:5, 1:5]
```

### Undifferentiated

```{r}
# load the gene expression matrix
undiff_mat <- readMM("data/processed-data/undifferentiated/counts_unfiltered/cells_x_genes.mtx")

# load barcode and gene information
undiff_barcodes <- readLines("data/processed-data/undifferentiated/counts_unfiltered/cells_x_genes.barcodes.txt")
undiff_genes <- readLines("data/processed-data/undifferentiated/counts_unfiltered/cells_x_genes.genes.txt")
undiff_gene_names <- readLines("data/processed-data/undifferentiated/counts_unfiltered/cells_x_genes.genes.names.txt")
```

```{r}
# Check the dimension and class of the gene expression matrix
dim(undiff_mat)
class(undiff_mat)

# check the lengths of the vectors containing information on barcodes, gene ID and names
length(undiff_barcodes)
length(undiff_genes)
length(undiff_gene_names)
```

```{r}
# transpose the gene expression matrix so that it is compatible to be turned into a SingleCellExperiment object
undiff_mat <- t(undiff_mat)

dim(undiff_mat)

rownames(undiff_mat) <- undiff_genes
colnames(undiff_mat) <- undiff_barcodes

# check the matrix
# undiff_mat[1:5, 1:5]

```

------------------------------------------------------------------------

## Create the SingleCellExperiment objects (Adipo 3d and Undifferentiated)

```{r}
adipo_sce <- SingleCellExperiment(assays = list(counts = adipo_mat))
undiff_sce <- SingleCellExperiment(assays = list(counts = undiff_mat))

# adding the experimental condition
adipo_sce$condition <- "Adipo_3d"
undiff_sce$condition <- "Undifferentiated"
```

kb count output —-\> Matrix Market file —-\> R sparse matrix —-\> SingleCellExperiment

```{r}
# Sanity check the SCE objects
# adipo3d
dim(adipo_mat)
length(adipo_barcodes)
length(adipo_genes)
dim(adipo_sce)

# undifferentiated
dim(undiff_mat)
length(undiff_barcodes)
length(undiff_genes)
dim(undiff_sce)

```

## Cell Calling

We will be using the DropletUtils for cell calling

```{r fig.height=6, fig.width=8}
# Adipo_3D

adipo_ranks <- barcodeRanks(counts(adipo_sce))


plot(adipo_ranks$rank, adipo_ranks$total, log = "xy", xlab = "Barcode rank", ylab = "Total UMI count", main = "Adipo 3d barcode rank")

o <- order(adipo_ranks$rank)

lines(adipo_ranks$rank[o], adipo_ranks$fitted[o])

abline(h = metadata(adipo_ranks)$knee, lty = 2)

abline(h = metadata(adipo_ranks)$inflection, lty = 2)

metadata(adipo_ranks)$knee
metadata(adipo_ranks)$inflection

```

```{r fig.height=6, fig.width=8}
# Undifferentiated

# calculate the barcode rank
undiff_ranks <- barcodeRanks(counts(undiff_sce))

# create the plot to visualize the knee and inflection
plot(undiff_ranks$rank, undiff_ranks$total, log = "xy", xlab = "Barcode rank", ylab = "Total UMI count", main = "Undiff barcode rank")

o <- order(undiff_ranks$rank)

lines(undiff_ranks$rank[o], undiff_ranks$fitted[o])

abline(h = metadata(undiff_ranks)$knee, lty = 2)

abline(h = metadata(undiff_ranks)$inflection, lty = 2)

# store the values as metadata
metadata(undiff_ranks)$knee
metadata(undiff_ranks)$inflection

```

> ***Note to self: Need to better understand what exactly is happening in the bar code rank and how to properly interpret the plot and terminologies such as knee and inflection***

```{r}
# dropping droplets with ambient RNA (will take some time)

adipo_emptydrops <- emptyDrops(
    counts(adipo_sce)
)
```

```{r}
head(adipo_emptydrops)
```

```{r}
table(adipo_emptydrops$Limited)
```

```{r}
sum(adipo_emptydrops$FDR <= 0.01, na.rm = TRUE)
```

> 8,884 barcodes have been statistically called as cell-containing droplets at FDR ≤ 1%.

Downstream processing so far,

```         
              788,172 barcodes
                     │
                     ▼
              barcode ranks
                     │
                     ▼
        ┌────────────────────────┐
        │ high-count population  │
        │                        │
        │   ↓                    │
        │ likely cells           │
        └────────────────────────┘
                     │
                emptyDrops()
                     │
                     ▼
                8,884 calls
                     │
                     ▼
            candidate cells
```

#### Check UMI distribution of the candidate cells

```{r}
adipo_called <- adipo_emptydrops$FDR <= 0.01

summary(adipo_emptydrops$Total[adipo_called])
```

```         
minimum called barcode = 101 UMIs
median = 8174 UMIs
maximum = 41425 UMIs
```

As we found earlier the knee was 8,168.916 UMIs and our **called cell population** has a median UMI of 8,174, which is a nice sanity check meaning the statistically called population is concentrated around the high-count region identified by the **barcode-rank analysis**.

```{r}
adipo_called <- !is.na(adipo_emptydrops$FDR) &
                adipo_emptydrops$FDR <= 0.01

sum(adipo_called)

quantile(
    adipo_emptydrops$Total[adipo_called],
    probs = c(0, 0.01, 0.05, 0.25, 0.5, 0.75, 0.95, 0.99, 1)
)
```

8,789 candidate non-empty/cell-containing droplets

------------------------------------------------------------------------

## Quality Control

```{r}
adipo_sce_called <- adipo_sce[, adipo_called]
dim(adipo_sce_called)

```

```{r}
adipo_qc <- perCellQCMetrics(
    adipo_sce_called
)
```

```{r}
#adipo_qc$sum
#adipo_qc$detected
```

sum = total UMI counts per cell

detected = number of genes detected per cell

```{r}
colData(adipo_sce_called)$total_counts <- adipo_qc$sum
colData(adipo_sce_called)$detected_genes <- adipo_qc$detected
```

```{r}
head(colData(adipo_sce_called))
```

```{r fig.height=6, fig.width=8}
hist(
    adipo_sce_called$total_counts,
    breaks = 100,
    main = "Adipo 3d: total UMI counts per called cell",
    xlab = "Total UMI counts"
)
```

```{r fig.height=6, fig.width=8}
hist(
    adipo_sce_called$detected_genes,
    breaks = 100,
    main = "Adipo 3d: detected genes per called cell",
    xlab = "Number of detected genes"
)
```

```{r}
summary(adipo_sce_called$total_counts)
summary(adipo_sce_called$detected_genes)
```

#### Adding Mitochondria Gene Names and calculating MT gene percentage

```{r}
rowData(adipo_sce)$gene_name <- adipo_gene_names
head(rowData(adipo_sce))
```

```{r}
head(rowData(adipo_sce)$gene_name, 20)
```

```{r}
sum(grepl("^MT-", rowData(adipo_sce)$gene_name))
```

```{r}
is_mito <- grepl(
    "^MT-",
    rowData(adipo_sce)$gene_name
)

adipo_qc <- perCellQCMetrics(
    adipo_sce_called,
    subsets = list(
        mito = is_mito
    )
)

colData(adipo_sce_called)$mito_percent <- adipo_qc$subsets_mito_percent

summary(adipo_sce_called$mito_percent)
```

```{r}
hist(
    adipo_sce_called$mito_percent,
    breaks = 100,
    main = "Adipo 3d: mitochondrial percentage",
    xlab = "Mitochondrial counts (%)"
)
```

```{r}
plot(
    adipo_sce_called$total_counts,
    adipo_sce_called$mito_percent,
    log = "x",
    pch = 16,
    cex = 0.5,
    xlab = "Total UMI counts",
    ylab = "Mitochondrial counts (%)",
    main = "Adipo 3d: UMI counts vs mitochondrial percentage"
)
```

#### Identifying QC outliers

```{r}
low_umi <- isOutlier(
    adipo_sce_called$total_counts,
    log = TRUE,
    type = "lower",
    nmads = 3
)

low_genes <- isOutlier(
    adipo_sce_called$detected_genes,
    log = TRUE,
    type = "lower",
    nmads = 3
)


high_mito <- isOutlier(
    adipo_sce_called$mito_percent,
    type = "higher",
    nmads = 3
)
```

```{r}
table(low_umi)
table(low_genes)
table(high_mito)
```

```{r}
# tells us how many cells would be flagged by at least one QC criterion
table(low_umi | low_genes | high_mito)
```

```{r}
table(
    low_umi,
    low_genes
)
```

```{r}
table(
    low_umi,
    high_mito
)
```

```{r}
table(
    low_genes,
    high_mito
)
```

```{r}
attr(low_umi, "thresholds")

attr(low_genes, "thresholds")

attr(high_mito, "thresholds")
```

Three groups of cell present in the dataset

```         
                     8,789 called droplets
                             │
          ┌──────────────────┼──────────────────┐
          │                  │                  │
          ▼                  ▼                  ▼
     main population     low-RNA cells      high-mito cells
       ~5,400             ~3,100              ~260
```

```{r}
plot(
    adipo_sce_called$total_counts,
    adipo_sce_called$detected_genes,
    log = "xy",
    pch = 16,
    cex = 0.5,
    xlab = "Total UMI counts",
    ylab = "Detected genes",
    main = "Adipo 3d: UMI counts vs detected genes"
)
```

#### Filtering the original *SingleCellExperiment* object to retain only high quality cells

Continuing with cell level QC to only retain high quality cells in the final *SingleCellExperiment* object.

```{r}
colData(adipo_sce_called)$low_umi <- low_umi
colData(adipo_sce_called)$low_genes <- low_genes
colData(adipo_sce_called)$high_mito <- high_mito

# criterion for high quality cells
adipo_keep <- !low_umi & !low_genes & !high_mito

# create a new SCE object by filtering the original SCE object to only retain high quality cells
adipo_sce_filtered <- adipo_sce_called[, adipo_keep]

# check the new SCE object
dim(adipo_sce_filtered)
```

> Now the next step is to find droplets which contains more than one cell and remove them

```{r}
adipo_sce_filtered <- scDblFinder(adipo_sce_filtered)

table(adipo_sce_filtered$scDblFinder.class)

table(adipo_sce_filtered$scDblFinder.class,
      useNA = "ifany")
```

```{r}
# removing the droplets containing more than one cell
adipo_sce_clean <- adipo_sce_filtered[
    ,
    adipo_sce_filtered$scDblFinder.class == "singlet"
]

dim(adipo_sce_clean)
```